\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-08-10 paper/ms.tex
comma after e.g. and i.e.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist
\citep[e.g.][]{feroz2009multinest, handley2015polychord}.
Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
is an alternative to NS for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm, and evolves one or more particles in the parameter space, along with
an integer index variable $j$ for each particle,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels},
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose log likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
each of which compresses the distribution by a factor of about $e \approx 2.71818$
(see \citet{brewer2011diffusive} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by
\citet{brewer2011diffusive}.

The mixture of constrained priors tends to be easier to sample than the
posterior, as the prior is always a mixture component, allowing the
MCMC chain to mix between different modes in some circumstances. The marginal
likelihood estimate can also be more accurate than standard MCMC-based NS
\citep{brewer2011diffusive}, as less information is discarded.
DNS has been applied several times in astrophysics
\citep[e.g.][]{pancoast2014modelling, huppenkothen2015dissecting,
brewer2015fast}
and was recently used in a biological application
\citep{dybowski2015single}.

@@

In DNS, the target distribution is not the posterior but rather
the joint distribution in Equation~\ref{eqn:target_distribution}.
Moves of $\theta$ are done keeping $j$ fixed, so we only need
to consider the Metropolis acceptance probability for fixed $j$,
i.e. with respect to a single constrained prior like
Equation~\ref{eqn:constrained_prior}.
Hence, the appropriate acceptance probability
for a proposed move from $\theta$ to $\theta'$ is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > \ell_j\right)
\right]
\label{eqn:log_hastings}
\end{align}
where $\ell_j$ is the likelihood threshold for the current level $j$.
There are also moves that propose a change to $j$ while keeping
$\params$ fixed, but the details are less relevant to the user.

@@

The \proglang{Python}
packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
and \pkg{Cython} \citep{cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:
\begin{CodeChunk}
\begin{CodeInput}
wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
tar xvzf 0.1.2.tar.gz
mv DNest4-0.1.2 DNest4
cd DNest4/code
make
cd ../python
python setup.py install
\end{CodeInput}
\end{CodeChunk}

@@

\section{Output files}
The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e. running the MCMC chain, building
levels, and then exploring all the levels). It creates three text output files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.

@@

The third output file, {\tt levels.txt}, contains information about the levels
that were built during the run. The first column has estimates of the $\log(X)$
values of the levels, i.e. how compressed they are relative to the prior, in
units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
value encloses $\exp(-1.02) \approx 36.1\%$ of the prior mass.

@@

\subsection{Save interval}
The save interval controls how often \dnest~writes a model to the output
files; what is usually called ``thinning''. Saving more frequently
(i.e. a smaller save interval) is usually better. However, this can result
in
big output files if your model prints a lot of parameters to {\tt sample.txt}.
causing the postprocessing to take a long time and/or a lot of RAM.
A default suggestion, used in the example, is to set the save interval to the
same value as the new level interval.

@@

As described by \citet{skilling2006nested}, there is no guarantee in principle that
another peak might have appeared had a run continued for longer. These
{\em phase changes} are more common in statistical mechanics problems than
in data analysis problems, but can appear in the latter
\citep[e.g.][]{brewer2014inference, brewer2015fast}.

@@

    if(which == 0)
    {
        // log_H takes care of the prior ratio
        // i.e. log_H = log(pi(theta')/pi(theta))
        log_H -= -0.5*pow(m/1E3, 2);

@@

To access the component parameters, the \code{RJObject} member function
\code{get_components} is used. This was used in the above functions, but more
typically it is needed in the log likelihood.
The member function \code{get_components} returns (by const reference) a
\code{std::vector} of \code{std::vector}s of \code{double}s. For example,
parameter two of component zero is:
\begin{CodeChunk}
\begin{CodeInput}
gaussians.get_components()[0][2]
\end{CodeInput}
\end{CodeChunk}
The order of $\{(\mu_j, \sigma_j, w_j)\}$
(i.e. which one is parameter 0, 1, and 2)
is determined by the conditional prior class, and is
explained in the following subsection.

@@

\subsection{Conditional priors}
The \code{RJObject} class is used to manage the components. An additional
class is needed to define and manage the hyperparameters $\alpha$ (i.e.
define how they are generated from the prior, how they are proposed, and
so on). The iid
conditional prior $p(x_i | \alpha)$ is also defined by this extra class.
In the example, the class is called \code{MyConditionalPrior} and
is inherited from an abstract base class \code{DNest4::ConditionalPrior}
which insists that certain member functions be specified.

@@

The functions \code{to\_uniform} and \code{from\_uniform} must implement
the cumulative distribution function (CDF) of the conditional
prior and its inverse. i.e. \code{from\_uniform}, if the input vector contains
iid draws from Uniform(0, 1), these should be modified to become draws from
$p(x|\alpha)$, and \code{to\_uniform} should be the inverse of
\code{from\_uniform}.
Both of these functions modify the vector argument in-place.
For the example, we again used an external class to define these functions
for the Laplace distribution:

@@

This approach is rarely usable in practice because the probability of
generating a dataset that matches the observed one is extremely low.
To work around this, we replace the proposition
$D = D_{\rm observed}$ with a logically
weaker one (i.e. one that is implied by $D = D_{\rm observed}$ but does not
imply it).
The weaker proposition is defined using
a discrepancy function
$\rho$, which measures how different a simulated dataset $D$ is from
the real one $D_{\rm observed}$:
\begin{align}
\rho\left(D; D_{\rm observed}\right).
\end{align}
The discrepancy function should take a minimum value of zero when
$D = D_{\rm observed}$.
The analysis then proceeds by Monte Carlo sampling the
joint posterior for $\theta$ and $D$
conditional on the logically weaker proposition
\begin{align}
\rho\left(D; D_{\rm observed}\right) < \epsilon
\end{align}
where $\epsilon$ is some small number. With this proposition, the
rejection rate will still
typically be very high, but lower than with $D = D_{\rm observed}$.
Typically, $\rho$ is defined by introducing a few {\em summary statistics}
$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
information in the data, and then computing the Euclidian distance
between the summary statistics of $D$ and $D_{\rm observed}$.
That is,
\begin{align}
\rho(D; D_{\rm observed}) &=
\sqrt{\sum_i \left[s_i(D) - s_i(D_{\rm observed})\right]^2}.
\end{align}

The main challenges associated with ABC are:
\begin{enumerate}[(i)]
\item The choice of discrepancy function $\rho(D; D_{\rm observed})$, which
may involve a choice of summary statistics,
\item How to choose the value of $\epsilon$, or how to change it as a run
progresses,
\item How to make algorithms more efficient than rejection sampling.
\end{enumerate}
Challenge (i) is Bayesian in nature, i.e. it relates to the very
definition of the posterior distribution itself.
On the other hand, challenges (ii) and (iii) are about the computational
implementation. Most ABC analyses are done using Sequential Monte Carlo
\citep[SMC; ][]{del2012adaptive}, another family of algorithms closely related
to Nested Sampling and annealing in the sense that they work with
sequences of probability distributions.

>>>>

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist
\citep[e.g.,][]{feroz2009multinest, handley2015polychord}.
Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
is an alternative to NS for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm, and evolves one or more particles in the parameter space, along with
an integer index variable $j$ for each particle,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels},
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose log likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
each of which compresses the distribution by a factor of about $e \approx 2.71818$
(see \citet{brewer2011diffusive} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by
\citet{brewer2011diffusive}.

The mixture of constrained priors tends to be easier to sample than the
posterior, as the prior is always a mixture component, allowing the
MCMC chain to mix between different modes in some circumstances. The marginal
likelihood estimate can also be more accurate than standard MCMC-based NS
\citep{brewer2011diffusive}, as less information is discarded.
DNS has been applied several times in astrophysics
\citep[e.g.,][]{pancoast2014modelling, huppenkothen2015dissecting,
brewer2015fast}
and was recently used in a biological application
\citep{dybowski2015single}.

@@

In DNS, the target distribution is not the posterior but rather
the joint distribution in Equation~\ref{eqn:target_distribution}.
Moves of $\theta$ are done keeping $j$ fixed, so we only need
to consider the Metropolis acceptance probability for fixed $j$,
i.e., with respect to a single constrained prior like
Equation~\ref{eqn:constrained_prior}.
Hence, the appropriate acceptance probability
for a proposed move from $\theta$ to $\theta'$ is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > \ell_j\right)
\right]
\label{eqn:log_hastings}
\end{align}
where $\ell_j$ is the likelihood threshold for the current level $j$.
There are also moves that propose a change to $j$ while keeping
$\params$ fixed, but the details are less relevant to the user.

@@

The \proglang{Python}
packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
and \pkg{Cython} \citep{cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:
\begin{CodeChunk}
\begin{CodeInput}
> wget https://github.com/eggplantbren/DNest4/archive/0.1.2.tar.gz
> tar xvzf 0.1.2.tar.gz
> mv DNest4-0.1.2 DNest4
> cd DNest4/code
> make
> cd ../python
> python setup.py install
\end{CodeInput}
\end{CodeChunk}

@@

\section{Output files}
The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e., running the MCMC chain, building
levels, and then exploring all the levels). It creates three text output files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.

@@

The third output file, {\tt levels.txt}, contains information about the levels
that were built during the run. The first column has estimates of the $\log(X)$
values of the levels, i.e., how compressed they are relative to the prior, in
units of nats. For example, if a level has $\log(X) = -1.02$, its likelihood
value encloses $\exp(-1.02) \approx 36.1\%$ of the prior mass.

@@

\subsection{Save interval}
The save interval controls how often \dnest~writes a model to the output
files; what is usually called ``thinning''. Saving more frequently
(i.e., a smaller save interval) is usually better. However, this can result
in
big output files if your model prints a lot of parameters to {\tt sample.txt}.
causing the postprocessing to take a long time and/or a lot of RAM.
A default suggestion, used in the example, is to set the save interval to the
same value as the new level interval.

@@

As described by \citet{skilling2006nested}, there is no guarantee in principle that
another peak might have appeared had a run continued for longer. These
{\em phase changes} are more common in statistical mechanics problems than
in data analysis problems, but can appear in the latter
\citep[e.g.,][]{brewer2014inference, brewer2015fast}.

@@

    if(which == 0)
    {
        // log_H takes care of the prior ratio
        // i.e., log_H = log(pi(theta')/pi(theta))
        log_H -= -0.5*pow(m/1E3, 2);

@@

To access the component parameters, the \code{RJObject} member function
\code{get_components} is used. This was used in the above functions, but more
typically it is needed in the log likelihood.
The member function \code{get_components} returns (by const reference) a
\code{std::vector} of \code{std::vector}s of \code{double}s. For example,
parameter two of component zero is:
\begin{CodeChunk}
\begin{CodeInput}
gaussians.get_components()[0][2]
\end{CodeInput}
\end{CodeChunk}
The order of $\{(\mu_j, \sigma_j, w_j)\}$
(i.e., which one is parameter 0, 1, and 2)
is determined by the conditional prior class, and is
explained in the following subsection.

@@

\subsection{Conditional priors}
The \code{RJObject} class is used to manage the components. An additional
class is needed to define and manage the hyperparameters $\alpha$ (i.e.,
define how they are generated from the prior, how they are proposed, and
so on). The iid
conditional prior $p(x_i | \alpha)$ is also defined by this extra class.
In the example, the class is called \code{MyConditionalPrior} and
is inherited from an abstract base class \code{DNest4::ConditionalPrior}
which insists that certain member functions be specified.

@@

The functions \code{to\_uniform} and \code{from\_uniform} must implement
the cumulative distribution function (CDF) of the conditional
prior and its inverse. i.e., \code{from\_uniform}, if the input vector contains
iid draws from Uniform(0, 1), these should be modified to become draws from
$p(x|\alpha)$, and \code{to\_uniform} should be the inverse of
\code{from\_uniform}.
Both of these functions modify the vector argument in-place.
For the example, we again used an external class to define these functions
for the Laplace distribution:

@@

This approach is rarely usable in practice because the probability of
generating a dataset that matches the observed one is extremely low.
To work around this, we replace the proposition
$D = D_{\rm observed}$ with a logically
weaker one (i.e., one that is implied by $D = D_{\rm observed}$ but does not
imply it).
The weaker proposition is defined using
a discrepancy function
$\rho$, which measures how different a simulated dataset $D$ is from
the real one $D_{\rm observed}$:
\begin{align}
\rho\left(D; D_{\rm observed}\right).
\end{align}
The discrepancy function should take a minimum value of zero when
$D = D_{\rm observed}$.
The analysis then proceeds by Monte Carlo sampling the
joint posterior for $\theta$ and $D$
conditional on the logically weaker proposition
\begin{align}
\rho\left(D; D_{\rm observed}\right) < \epsilon
\end{align}
where $\epsilon$ is some small number. With this proposition, the
rejection rate will still
typically be very high, but lower than with $D = D_{\rm observed}$.
Typically, $\rho$ is defined by introducing a few {\em summary statistics}
$s_1(D), s_2(D), ..., s_n(D)$, which hopefully capture most of the relevant
information in the data, and then computing the Euclidian distance
between the summary statistics of $D$ and $D_{\rm observed}$.
That is,
\begin{align}
\rho(D; D_{\rm observed}) &=
\sqrt{\sum_i \left[s_i(D) - s_i(D_{\rm observed})\right]^2}.
\end{align}

The main challenges associated with ABC are:
\begin{enumerate}[(i)]
\item The choice of discrepancy function $\rho(D; D_{\rm observed})$, which
may involve a choice of summary statistics,
\item How to choose the value of $\epsilon$, or how to change it as a run
progresses,
\item How to make algorithms more efficient than rejection sampling.
\end{enumerate}
Challenge (i) is Bayesian in nature, i.e., it relates to the very
definition of the posterior distribution itself.
On the other hand, challenges (ii) and (iii) are about the computational
implementation. Most ABC analyses are done using Sequential Monte Carlo
\citep[SMC; ][]{del2012adaptive}, another family of algorithms closely related
to Nested Sampling and annealing in the sense that they work with
sequences of probability distributions.
